Supplemental Materials: \\ \emph{Active-contours driven registration method for structure-informed segmentation of diffusion MR images}

### 2015-04-29 Publications/2015-NeuroImage/04-methods.tex

% -*- root: 00-main.tex -*-
\section{Methods}
\label{sec:methods}
%
Let $\gammaset_R = \{\Gamma_m: m \in \mathbb{N}, m \leq N_c\}$ be the set of $N_c$ surfaces
  extracted from the undistorted \gls*{t1} image (the reference space $R$).
We reformulate the segmentation of the distorted \gls*{dmri} images (the moving space $M$)
  as a registration problem where we look for an underlying deformation field $U$, such that 
  the structures in $R$ defined by $\gammaset_R$ optimally align with their corresponding
  structures in $M$:

  \begin{align}
  U\colon R \subset \mathbb{R}^n &\to M \subset \mathbb{R}^n \notag\\
  \vec{r} &\mapsto \vec{r}' =\vec{r}+\vec{u}(\vec{r}),
  \label{eq:transform}
  \end{align}
%
  where $\vec{r}$ denotes a position in $R$, $\vec{r}'$ is
  its corresponding location in $M$, and $n$ the dimensionality of images.
Finally, $\vec{u} = \vec{u}(\vec{r})$ is the displacement of every point with respect
  to the reference domain.

\subsection{Cost-function derivation}\label{sec:methods_map}
%
In a Bayesian framework for registration \citep{wyatt_map_2003,pohl_bayesian_2006,gass_simultaneous_2014},
  the mappings $U$ in \eqref{eq:transform} are
  evaluated based on their posterior probability given the observed data
  $M$.
Let $\omegaset = \{\Omega_l: l \in \mathbb{N}, l \leq N_L\}$ be the set of $N_L$ competing regions in
  which $M$ is partitioned by the projection of $\gammaset_R$.
Using the Bayes' rule, the posterior likelihood is computed as:

  \begin{equation}
  P(U \mid M, \omegaset ) = \frac{P(M \mid U,\omegaset )\, P(U)}{P(M)},
  \label{eq:bayes_rule}
  \end{equation}
%
  where $P(M \mid U,\omegaset)$ is the data-likelihood.
As $\omegaset$ are mapped by $U$, we simplify
  $P(U, \omegaset) = P(U) \implies P(M \mid U,\omegaset) = P(M \mid U)$.

The best estimate $\hat{U}$ then fulfills the maximum a posteriori criterium
  \citep{bishop_pattern_2006}, and aligns $\gammaset_R$ into $M$.
First, we assume independence between pixels, and thus break down the
  global data likelihood into a product of pixel-wise conditional probabilities:

  \begin{equation}
  P(M \mid U) = \underset{l}{\prod} \underset{\vec{r}\in \Omega_l}{\prod}
    P\left( \vec{f}' \mid U \right),
  \label{eq:bayes_aposteriori}
  \end{equation}
%
  where $\vec{f}' = M(\vec{r}')$ is the feature vector at the displaced
  position $\vec{r}'$ \eqref{eq:transform} in the moving image.
For convenience, and because it has been shown to be an appropriate approximation
  \citep{leemput_automated_1999,cuadra_comparison_2005}, we introduce two assumptions for each
  region $\Omega_l$:
  1) the features are i.i.d.; and 2) they can be modeled by multivariate normal
  distributions with parameters $\lbrace \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_{l} \rbrace$
  for each region $\Omega_l$:

 	\begin{align}
  P( M \mid U) &= \underset{l}{\prod} \underset{\vec{r} \in \Omega_l}{\prod}
  \mathcal{N} ( \vec{f} \mid \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_{l} ) = % \notag\\ &=
  \underset{l}{\prod} \underset{\vec{r} \in \Omega_l}{\prod} \frac{1}{ \sqrt{(2\pi)^{C}\,\left|\boldsymbol{\Sigma}_{l}\right|}}\,{e^{\left(-\frac{1}{2}
  \mdist{f'}{l} \right)}},
  \label{eq:pdf}
  \end{align}
%
  using $\mdist{f}{l}$ to denote the squared \emph{Mahalanobis distance} of $\vec{f}$ with respect
  to the descriptors of region $l$ as
  $\mdist{f}{l} = (\vec{f} - \boldsymbol{\mu}_l)^T \, {\boldsymbol{\Sigma}_l}^{-1} \, (\vec{f} - \boldsymbol{\mu}_l)$.
$C$ is the number of channels comprised in the image $M$.
In fact, the projection of $\gammaset_R$ onto $M$ is an implicit segmentation model, for which
  the covariance matrix $\boldsymbol{\Sigma}_l$ of each region is minimized.

@@

  \begin{align}
  P(U) = \underset{\vec{r}}{\prod}\, p(\vec{u}) &=
  \underset{\vec{r}}{\prod}\, p_0(\vec{u}) \, p_1(\vec{u}), \text{ with} \\
  p_0(\vec{u}) &= \mathcal{N}( \vec{u} \mid 0, \mathbf{A}^{-1}), \notag\\
  p_1(\vec{u}) &= \mathcal{N}(  \nabla \vec{u} \mid 0, \mathbf{B}^{-1}),
  \label{eq:thikonov}
  \end{align}
%
  imposing that the distortion and its gradient have zero
  mean and variance governed by the matrices $\mathbf{A}$ and $\mathbf{B}$.
Finally, the maximum a posteriori problem is adapted into a variational one where we look for
  the minimum of an energy functional, by applying $E(\vec{u}) = -\log P(U)$:

  \begin{align}
  E(\vec{u}) &= -\log \underset{l}{\prod}
  \underset{\vec{r} \in \Omega_l}{\prod}
  \mathcal{N} \left( \vec{f}' \mid \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_l \right)\,p_0( \vec{u})\,p_1( \vec{u}) = \notag\\ &=
  \const + \underset{l}{\sum} \int_{\Omega_l} \mdist{f'}{l} \,d\vec{r} \, + % \notag\\ &+
  \int_{\Omega} \frac12 \left[ \vec{u}^T \mathbf{A} \vec{u} + (\nabla \vec{u})^T \mathbf{B} (\nabla \vec{u}) \right] \,d\vec{r}.
  \label{eq:energy}
  \end{align}
%
This expression is dual of the Mumford-Shah functional corresponding
  to a framework of \acrlong*{acwe} \citep{chan_active_2001}
  with the corresponding anisotropic regularization term of \cite{nagel_investigation_1986}.

@@

\begin{figure}
	\includegraphics[width=\linewidth]{figure01}
	\caption{The active contours are defined as the interfacing surfaces of the competing
	  \glspl{roi} $\Omega_i$.
	They are represented in green and dark blue colors in this close-up.
	They iteratively evolve following their inward normals $\hat{n}_i$ at each vertex
	  $\vec{v}_i$ of the mesh.
	The gradient speeds $\bar{s}_i$ drive registration, and are computed as the disparity of data
    energies with respect to the two limiting regions of $M(\vec{v}_i)$, the features of the image
    $M$ in the location of vertex $\vec{v}_i$ (see \suppl{equation SM5}).
	In this figure, $\bar{s}_1$ is written in the lower
	  box, with $\Omega_{wm}$ being the inner limiting region, $\Omega_{gm}$ the outer region, and
    $w_{0,1}$ the relative area associated to vertex $\vec{v}_1$ with respect to
    the total area of surface $\Gamma_0$.
	% Finally, every $\bar{s}_i$ is associated with the control points $\vec{u}_k$ through
	%   the corresponding weights $\psi_k(\vec{v}_i)$ by Eq. \eqref{eq:gradient_wshape}.
	}\label{fig:method}
\end{figure}

\paragraph*{Deformation model}\label{sec:deformation_model}
Since the vertices of the surfaces $\{\vec{v}_i: \vec{v}_i \subset \gammaset \}_{i=1 \ldots N_v}$
  likely lay off-grid, it is required to derive $\vec{u}_i = \vec{u}(\vec{v}_i)$ from a discrete set of parameters
  $\{\vec{u}_k\}_{k=1 \ldots K}$.
Densification is achieved through a set of associated basis functions $\psi_k$ \eqref{eq:nodes_tfm}.
In our implementation, $\psi_k$ is chosen to be a tensor-product B-Spline kernel
  of degree 3.

@@

\paragraph*{Optimization}
\label{sec:gradient_descent}
To find the minimum of the energy functional \eqref{eq:energy},
  we propose a gradient-descent approach with respect to the underlying
  deformation field through the following \gls*{pde}:

  \begin{equation}
  \frac{\partial \vec{u}(\vec{r},t)}{\partial t} \propto - \frac{\partial E(\vec{u})}{\partial \vec{u}_k},
  \label{eq:general_gradient_descent}
  \end{equation}
%
  with $t$ being an artificial time parameter of the contour
  evolution, and $\vec{u}_k$ the parameters supporting the estimate
  $\hat{U}$ of the transformation at the current time point.
Let us assume that anisotropy is aligned with the imaging axes to simplify
  \eqref{eq:energy} into expression \eqref{eq:app_energy} of \ref{app:reg_term}
  and compute its derivative \eqref{eq:general_gradient_descent}:

  \begin{align}
  \frac{\partial E(\vec{u})}{\partial \vec{u}_k} &=
  \frac{ \partial }{\partial \vec{u}_k} \Big\{
  \underset{l}{\sum} \int_{\Omega_l} \mdist{f'}{l} \,d\vec{r} \, + % \notag\\ &+
  \int_{\Omega} \frac12 [ \boldsymbol{\alpha} \cdot \vec{u}^{\circ2}
  + \boldsymbol{\beta} \cdot (\nabla \vec{u})^{\circ2} ] \,d\vec{r}
  \Big\},
  \label{eq:gradient_descent}
  \end{align}
%
  where $\vec{u}^{\circ2} = \vec{u}^T \cdot \vec{u}$.
Then, the data and regularization terms are split and discretized to compute their
  derivatives.
The computation of the explicit shape gradients at each $\vec{v}_i'$ is described in \autoref{fig:method}.
Then, \eqref{eq:gradient_descent} can be reformulated as (see \suppl{equations SM6-SM12}):

@@

Finally, to descend this gradient we establish a semi-implicit Euler scheme (see \suppl{section S1.3}),
  with a step size parameter $\delta$, that we solve in the spectral domain as follows:

  \begin{align}
  \vec{u}_k^{t+1} = \mathcal{F}^{-1}\left\{ \frac{\mathcal{F}\{\delta^{-1} \, \vec{u}_k^t - \vec{g}_k\} }%
                  {\mathcal{F}\{(\delta^{-1} + \boldsymbol{\alpha})\, I - \boldsymbol{\beta}\Delta\}} \right\},
  \label{eq:update_equation}
  \end{align}
  where $I$ denotes the identity operator.

\paragraph*{Implementation details, settings and convergence}
\label{sec:conv_report}
The \regseg{} tool includes a multiresolution strategy on the free-form deformation field.
Registration pyramids are created by setting the spacing between control points of the B-Spline basis
  functions for each level of the multiresolution.
This implementation detail, along with other features (as the sparse matrix approach
  to fast interpolation) and the main parameters 
  (such as $\delta$, $\boldsymbol{\alpha}$, $\boldsymbol{\beta}$, the B-Spline grid resolutions,
  target image smoothing, etc.) are discussed in the \suppl{section S1}.
The actual choices of parameter settings are publicly distributed with the source code of the experiments.
These final settings were manually obtained based on the feedback of post-registration convergence
  reports (as the one found in the \suppl{section S1.3}).
We release \regseg{} along with the tool for generating the convergence report.

\subsection{Evaluation protocol}\label{sec:evaluation_protocol}
In order to assess the performance of \regseg{}, we defined a general
  evaluation protocol:
1) Extraction of the set of undistorted surfaces $\gammaset_R$;
2) computation of a ground-truth field of displacements $U_{true}$ that is applied to
  generate warped images ($M$) to segment;
3) execution of \regseg{} with $\gammaset_R$ and the warped data as inputs; and
4) visual assessment and computation of error metrics.
How this protocol is adapted to the simulated phantoms and real data is covered in the
  following sections.

\subsection{Simulated phantoms}\label{sec:digital_phantoms}
The workflow to simulate the digital phantoms and assess
  \regseg{} on them is presented in \autoref{fig:evphantoms}.
A set of four binary objects (called ``box'', ``ball'', ``L'',
  and ``gyrus'') are generated combining binarization of
  analytical shapes and mathematical morphology.
The reference surfaces $\gammaset_R$ are extracted from the binary shapes
  using \emph{FreeSurfer} tools \citep{fischl_freesurfer_2012}.
The ground-truth distortion is generated with a chain of two displacement 
  fields supported by grids of B-Spline basis functions.
The coefficients of the basis functions are generated randomly for
  both levels in their three dimensions.
The three components of the displacements $\vec{u} = (u_d)$ 
  are bounded above by 40\% of the separation between control points
  at each level, to obtain diffeomorphic transforms
  after concatenation \citep{rueckert_diffeomorphic_2006}.
The first deformation field is intended to generate large warpings,
  with control points separated by 50.50mm along the three dimensions
  ($u_d\leq$ 20.20mm).
With the second warping, we look for a resulting field with smoothness
  close to the found in a typical distortion field of \gls*{dmri} data
  \citep{irfanoglu_susceptibility_2011}.
Therefore, the control points are separated 25.25mm ($u_d\leq$ 10.10mm).
Once the ground-truth deformation is generated, the original surfaces
  are warped interpolating the displacements field at each vertex.
The warped surfaces $\gammaset_\text{true}$ are binarized to generate tissue fractions
  in low (\isores[mm]{2.0}) and high (\isores[mm]{1.0}) resolution.
Using a \gls*{mr} simulator \citep{caruyer_phantomas_2014}, we synthesize
  \gls*{t1} (TE/TR= 10/1500ms) and \gls*{t2} images (TE/TR= 90/5000ms)
  corresponding to each phantom type, with two resolutions each
  (1.0mm and 2.0mm isotropic).
The field of view at both resolutions is \isores[mm]{100}.
Then, \regseg{} is applied to map $\gammaset_R$ into the warped phantoms to
  obtain the registered surfaces ($\hat{\gammaset}_{test}$).
% Simulations were corrupted with rician noise for a \gls*{snr} of 300.0.
To quantify the misregistration error, we compute the Hausdorff distance between
 $\hat{\gammaset}_{test}$ and $\gammaset_\text{true}$, using \citep{commandeur_vtk_2011}.

\begin{figure*}
\includegraphics[width=\linewidth]{figure02}
\caption{The evaluation of \regseg{} on phantom data is performed with the presented instrumental workflow:
  1) The reference surfaces $\gammaset_R$ are triangularized meshes extracted from the four binary shapes (``box'', ``ball'', ``L'', ``gyrus'');
  2) A ground-truth displacement field is generated as described in \autoref{sec:digital_phantoms}, and applied to warp
      $\gammaset_R$, obtaining $\gammaset_\text{true}$;
  3) Once warped, $\gammaset_\text{true}$ are projected to the corresponding discrete 3D volume, and downsampled creating partial volume effects to two resolutions
     of \isores[mm]{2.0} and \isores[mm]{1.0}, producing sets of tissue fractions maps;
  4) The tissue fractions are fed to an \acrfull*{mr} simulator, generating \acrfull*{t1} and \acrfull*{t2} -like images at the
     two available resolutions;
  5) The \regseg{} is run, using the warped test images as multispectral moving-image and $\gammaset_R$ as shape-priors;
  6) Assessment of the agreement between the surfaces fitted with \regseg{} ($\hat{\gammaset}_{test}$) and $\gammaset_\text{true}$
     with automatically generated visual reports and computing the Hausdorff distance between corresponding surfaces.}\label{fig:evphantoms}
\end{figure*}

\subsection{Real datasets} %
\label{sec:human_connectome}
% 
The general experimental framework for the real datasets is presented in \autoref{fig:evworkflows}.
It extends our previous work \citep{esteban_simulationbased_2014} evaluating distortions
  on \gls*{dmri} phantoms.

\begin{figure*}
\includegraphics[width=\linewidth]{figure03}
\caption{Experimental workflow real data from the \acrfull*{hcp}.
  1) $\gammaset_R$ are extracted from the anatomical reference (\gls*{t1} image).
  2) To operate as ground truth, we generate a plausible-synthetic distortion $U_{true}$
    from the fieldmap using \eqref{eq:fieldmap}.
  3) The \gls*{dmri} data are warped using $U_{true}$ to reproduce the effects of real
    susceptibility-derived distortions.
  Target diffusion scalars (\gls*{fa} and \gls*{adc}) are computed on the distorted data and
    stacked to feed the multivariate input required by \regseg{}.
  4) the method is run, obtaining a $U_{test} = \hat{U}_{true}$, the estimation of
    the ground-truth deformation.
  5) Results are visually and quantitatively evaluated.}\label{fig:evworkflows}
\end{figure*}

\paragraph*{Data}
For the evaluation of \regseg{} on real \gls*{dmri} data of human brains,
  we collected 16 subjects from the \gls*{hcp} database.
The original acquisitions are released within ``unprocessed'' packages, and
  the ``minimally preprocessed'' are corrected for artifacts, brain-extracted
  and spatially normalized, along with some results of the standard processing
  pipeline of \emph{FreeSurfer}.
We refer the reader to \citep{essen_human_2012} for exact details about acquisition
  parameters, and \citep{glasser_minimal_2013} for the preprocessing issues.
The datasets comprehend a large set of images, containing \gls*{t1}, \gls*{t2} and
  multi-shell \gls*{dmri} images.

\begin{figure}
\includegraphics[width=\linewidth]{figure04}
\caption{Segmentation model defined by the homogeneous regions $\Omega_i$ in which
  the image to be segmented is partitioned by the theoretical projection 
  ($\gammaset_{true}$) of $\gammaset_R$ into $M$.
The plot represents a kernel density estimation of the location and spread of
  each region $\Omega_i$ in the bivariate feature space of $M$, comprising 
  the \gls*{fa} and \gls*{adc} maps.
On top and right margins the marginal distributions of $\Omega_i$ are also plotted for
  the \gls*{fa} and \gls*{adc}, respectively.
}\label{fig:model}
\end{figure}

>>>>

% -*- root: 00-main.tex -*-
\section{Methods}
\label{sec:methods}
Let $\gammaset_R = \{\Gamma_m: m \in \mathbb{N}, m \leq N_c\}$ be the set of $N_c$ surfaces
  extracted from the undistorted \gls*{t1} image (the reference space $R$).
We reformulate the segmentation of the distorted \gls*{dmri} images (the moving space $M$)
  as a registration problem where we search for an underlying deformation field $U$ such that 
  the structures in $R$ defined by $\gammaset_R$ align optimally with their corresponding
  structures in $M$:

  \begin{align}
  U\colon R \subset \mathbb{R}^n &\to M \subset \mathbb{R}^n \notag\\
  \vec{r} &\mapsto \vec{r}' =\vec{r}+\vec{u}(\vec{r}),
  \label{eq:transform}
  \end{align}
  where $\vec{r}$ denotes a position in $R$, $\vec{r}'$ is
  its corresponding location in $M$, and $n$ denotes the dimensionality of images.
Finally, $\vec{u} = \vec{u}(\vec{r})$ is the displacement of every point with respect
  to the reference domain.

\subsection{Cost-function derivation}\label{sec:methods_map}
In a Bayesian framework for registration \citep{wyatt_map_2003,pohl_bayesian_2006,gass_simultaneous_2014},
  the mappings $U$ in \eqref{eq:transform} are
  evaluated based on their posterior probability given the observed data
  $M$.
Let $\omegaset = \{\Omega_l: l \in \mathbb{N}, l \leq N_L\}$ be the set of $N_L$ competing regions in
  which $M$ is partitioned by the projection of $\gammaset_R$.
Using Bayes' rule, the posterior likelihood is computed as:

  \begin{equation}
  P(U \mid M, \omegaset ) = \frac{P(M \mid U,\omegaset )\, P(U)}{P(M)},
  \label{eq:bayes_rule}
  \end{equation}
  where $P(M \mid U,\omegaset)$ is the data likelihood.
Because $\omegaset$ are mapped by $U$, then we simplify
  $P(U, \omegaset) = P(U) \implies P(M \mid U,\omegaset) = P(M \mid U)$.

The best estimate $\hat{U}$ then satisfies the maximum a posteriori criterion
  \citep{bishop_pattern_2006} and it aligns $\gammaset_R$ into $M$.
First, we assume independence between pixels, and thus we break down the
  global data likelihood into a product of pixel-wise conditional probabilities:

  \begin{equation}
  P(M \mid U) = \underset{l}{\prod} \underset{\vec{r}\in \Omega_l}{\prod}
    P\left( \vec{f}' \mid U \right),
  \label{eq:bayes_aposteriori}
  \end{equation}
  where $\vec{f}' = M(\vec{r}')$ is the feature vector at the displaced
  position $\vec{r}'$ \eqref{eq:transform} in the moving image.
For convenience and because it has been shown to be an appropriate approximation
  \citep{leemput_automated_1999,cuadra_comparison_2005}, we introduce two assumptions for each
  region $\Omega_l$:
  1) the features are i.i.d.; and 2) they can be modeled by multivariate normal
  distributions with the parameters $\lbrace \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_{l} \rbrace$
  for each region $\Omega_l$:

  \begin{align}
  P( M \mid U) &= \underset{l}{\prod} \underset{\vec{r} \in \Omega_l}{\prod}
  \mathcal{N} ( \vec{f} \mid \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_{l} ) =   \underset{l}{\prod} \underset{\vec{r} \in \Omega_l}{\prod} \frac{1}{ \sqrt{(2\pi)^{C}\,\left|\boldsymbol{\Sigma}_{l}\right|}}\,{e^{\left(-\frac{1}{2}
  \mdist{f'}{l} \right)}},
  \label{eq:pdf}
  \end{align}
  using $\mdist{f}{l}$ to denote the squared \emph{Mahalanobis distance} of $\vec{f}$ with respect
  to the descriptors of region $l$ as
  $\mdist{f}{l} = (\vec{f} - \boldsymbol{\mu}_l)^T \, {\boldsymbol{\Sigma}_l}^{-1} \, (\vec{f} - \boldsymbol{\mu}_l)$.
$C$ is the number of channels in the image $M$.
In fact, the projection of $\gammaset_R$ onto $M$ is an implicit segmentation model, for which
  the covariance matrix $\boldsymbol{\Sigma}_l$ of each region is minimized.

@@

  \begin{align}
  P(U) = \underset{\vec{r}}{\prod}\, p(\vec{u}) &=
  \underset{\vec{r}}{\prod}\, p_0(\vec{u}) \, p_1(\vec{u}), \text{ with} \\
  p_0(\vec{u}) &= \mathcal{N}( \vec{u} \mid 0, \mathbf{A}^{-1}), \notag\\
  p_1(\vec{u}) &= \mathcal{N}(  \nabla \vec{u} \mid 0, \mathbf{B}^{-1}),
  \label{eq:thikonov}
  \end{align}
 which requires that the distortion and its gradient have zero
  mean and variance governed by the matrices $\mathbf{A}$ and $\mathbf{B}$.
Finally, the maximum a posteriori problem is adapted to a variational problem where we search for
  the minimum of an energy functional by applying $E(\vec{u}) = -\log P(U)$:

  \begin{align}
  E(\vec{u}) &= -\log \underset{l}{\prod}
  \underset{\vec{r} \in \Omega_l}{\prod}
  \mathcal{N} \left( \vec{f}' \mid \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_l \right)\,p_0( \vec{u})\,p_1( \vec{u}) = \notag\\ &=
  \const + \underset{l}{\sum} \int_{\Omega_l} \mdist{f'}{l} \,d\vec{r} \, +   \int_{\Omega} \frac12 \left[ \vec{u}^T \mathbf{A} \vec{u} + (\nabla \vec{u})^T \mathbf{B} (\nabla \vec{u}) \right] \,d\vec{r}.
  \label{eq:energy}
  \end{align}
This expression is the dual of the Mumford-Shah functional that corresponds
  to the framework of \acrlong*{acwe} \citep{chan_active_2001}
  with the corresponding anisotropic regularization term of \cite{nagel_investigation_1986}.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figure01}
  \caption{The active contours are defined as the interfacing surfaces of the competing
    \glspl{roi} $\Omega_i$, which
  are represented in green and dark blue in this close-up.
  They evolve iteratively following their inward normals $\hat{n}_i$ at each vertex
    $\vec{v}_i$ of the mesh.
  The gradient speeds $\bar{s}_i$ drive registration, which are computed as the disparity of the data
    energies with respect to the two limiting regions of $M(\vec{v}_i)$, the features of the image
    $M$ in the location of vertex $\vec{v}_i$ (see \suppl{equation SM5}).
  In this figure, $\bar{s}_1$ is written in the lower
    box, with $\Omega_{wm}$ in the inner limiting region, $\Omega_{gm}$ the outer region, and
    $w_{0,1}$ is the relative area associated with vertex $\vec{v}_1$ with respect to
    the total area of surface $\Gamma_0$.
      }\label{fig:method}
\end{figure}

\paragraph*{Deformation model}\label{sec:deformation_model}
Since the vertices of the surfaces $\{\vec{v}_i: \vec{v}_i \subset \gammaset \}_{i=1 \ldots N_v}$
  are probably located off-grid, it is necessary to derive $\vec{u}_i = \vec{u}(\vec{v}_i)$ from a discrete set of parameters
  $\{\vec{u}_k\}_{k=1 \ldots K}$.
Densification is achieved using a set of associated basis functions $\psi_k$ \eqref{eq:nodes_tfm}.
In our implementation, $\psi_k$ is selected as a tensor-product B-spline kernel
  of degree three.

@@

\paragraph*{Optimization}
\label{sec:gradient_descent}
To find the minimum of the energy functional \eqref{eq:energy},
  we propose a gradient-descent approach with respect to the underlying
  deformation field using the following \gls*{pde}:

  \begin{equation}
  \frac{\partial \vec{u}(\vec{r},t)}{\partial t} \propto - \frac{\partial E(\vec{u})}{\partial \vec{u}_k},
  \label{eq:general_gradient_descent}
  \end{equation}
  where $t$ is an artificial time parameter of the contour
  evolution and $\vec{u}_k$ are the parameters that support the estimate
  $\hat{U}$ of the transformation at the current time point.
Let us assume that the anisotropy is aligned with the imaging axes to simplify
  \eqref{eq:energy} as expression \eqref{eq:app_energy} in \ref{app:reg_term},
  and thus to compute its derivative \eqref{eq:general_gradient_descent}:

  \begin{align}
  \frac{\partial E(\vec{u})}{\partial \vec{u}_k} &=
  \frac{ \partial }{\partial \vec{u}_k} \Big\{
  \underset{l}{\sum} \int_{\Omega_l} \mdist{f'}{l} \,d\vec{r} \, +   \int_{\Omega} \frac12 [ \boldsymbol{\alpha} \cdot \vec{u}^{\circ2}
  + \boldsymbol{\beta} \cdot (\nabla \vec{u})^{\circ2} ] \,d\vec{r}
  \Big\},
  \label{eq:gradient_descent}
  \end{align}
  where $\vec{u}^{\circ2} = \vec{u}^T \cdot \vec{u}$.
Then, the data and regularization terms are split and discretized to compute their
  derivatives.
The computation of the explicit shape gradients at each $\vec{v}_i'$ is illustrated in \autoref{fig:method}.
Then, \eqref{eq:gradient_descent} can be reformulated as (see \suppl{equations SM6-SM12}):

@@

Finally, to descend this gradient, we establish a semi-implicit Euler scheme (see \suppl{section S1.3}),
  with a step size parameter $\delta$, which we solve in the spectral domain as follows:

  \begin{align}
  \vec{u}_k^{t+1} = \mathcal{F}^{-1}\left\{ \frac{\mathcal{F}\{\delta^{-1} \, \vec{u}_k^t - \vec{g}_k\} }                  {\mathcal{F}\{(\delta^{-1} + \boldsymbol{\alpha})\, I - \boldsymbol{\beta}\Delta\}} \right\},
  \label{eq:update_equation}
  \end{align}
  where $I$ denotes the identity operator.

\paragraph*{Implementation details, settings, and convergence}
\label{sec:conv_report}
The \regseg{} tool includes a multiresolution strategy on the free-form deformation field.
Registration pyramids are created by setting the spacing between the control points of the B-spline basis
  functions for each level of the multiresolution strategy.
The implementation details as well as other features (such as the sparse matrix approach
  to fast interpolation) and the main parameters 
  (such as $\delta$, $\boldsymbol{\alpha}$, $\boldsymbol{\beta}$, the B-spline grid resolutions,
 and target image smoothing) are discussed in \suppl{section S1}.
The actual choices of the parameter settings are publicly distributed with the source code for the experiments.
The final settings were obtained manually based on feedback from the post-registration convergence
  reports (such as that found in \suppl{section S1.3}).
We released \regseg{} along with the tool to generate the convergence report.

\subsection{Evaluation protocol}\label{sec:evaluation_protocol}
In order to assess the performance of \regseg{}, we defined the following general
  evaluation protocol:
1) Extract the set of undistorted surfaces $\gammaset_R$;
2) Compute a ground-truth field of displacements $U_{true}$, which is applied to
  generate warped images ($M$) for segmentation;
3) Execute \regseg{} with $\gammaset_R$ and use the warped data as inputs; and
4) Perform a visual assessment and compute the error metrics.
The adaptation of this protocol to the simulated phantoms and real data is explained in the
  following sections.

\subsection{Simulated phantoms}\label{sec:digital_phantoms}
The workflow required to simulate the digital phantoms and to assess the performance of
  \regseg{} with them is presented in \autoref{fig:evphantoms}.
A set of four binary objects (i.e., ``box'', ``ball'', ``L'',
  and ``gyrus'') was generated by combining the binarization of
  analytical shapes and mathematical morphology.
The reference surfaces $\gammaset_R$ were extracted from the binary shapes
  using \emph{FreeSurfer} tools \citep{fischl_freesurfer_2012}.
The ground-truth distortion was generated using a chain of two displacement 
  fields supported by grids of B-spline basis functions.
The coefficients of the basis functions were generated randomly for
  both levels in their three dimensions.
The three components of the displacements $\vec{u} = (u_d)$ 
  were bounded above by 40\% of the separation between the control points
  at each level to obtain diffeomorphic transforms
  after concatenation \citep{rueckert_diffeomorphic_2006}.
The first deformation field was applied to generate large warpings
  with control points separated by 50.50 mm in the three dimensions
  ($u_d\leq$ 20.20 mm).
With the second warping, we aimed to obtain a field with smoothness
  close to that found in a typical distortion field of \gls*{dmri} data
  \citep{irfanoglu_susceptibility_2011}.
Therefore, the control points were separated by 25.25 mm ($u_d\leq$ 10.10 mm).
After generating the ground-truth deformation, the original surfaces
  were warped by interpolating the displacements field at each vertex.
The warped surfaces $\gammaset_\text{true}$ were binarized to generate tissue fractions
  at low (\isores[mm]{2.0}) and high (\isores[mm]{1.0}) resolutions.
Using a \gls*{mr} simulator \citep{caruyer_phantomas_2014}, we synthesized
  \gls*{t1} (TE/TR= 10/1500 ms) and \gls*{t2} images (TE/TR= 90/5000 ms), which
  corresponded to each phantom type, with each at two resolutions 
  (1.0 mm and 2.0 mm isotropic).
The field of view at both resolutions was \isores[mm]{100}.
Next, \regseg{} was applied to map $\gammaset_R$ onto the warped phantoms to
  obtain the registered surfaces ($\hat{\gammaset}_{test}$).
To quantify the misregistration error, we computed the Hausdorff distance between
 $\hat{\gammaset}_{test}$ and $\gammaset_\text{true}$ using \citep{commandeur_vtk_2011}.

\begin{figure*}
\includegraphics[width=\linewidth]{figure02}
\caption{Evaluation of \regseg{} using phantom data according to the following instrumental workflow.
  1) The reference surfaces $\gammaset_R$ are triangularized meshes extracted from the four binary shapes (i.e., ``box'', ``ball'', ``L'', ``gyrus'').
  2) A ground-truth displacement field was generated as described in \autoref{sec:digital_phantoms}, and applied to warp
      $\gammaset_R$, thereby obtaining $\gammaset_\text{true}$.
  3) After being warped, $\gammaset_\text{true}$ were projected onto the corresponding discrete 3D volume and downsampled to create partial volume effects at two resolutions,
     i.e., \isores[mm]{2.0} and \isores[mm]{1.0}, thereby producing sets of tissue fractions maps.
  4) The tissue fractions were fed into an \acrfull*{mr} simulator, which generated \acrfull*{t1} and \acrfull*{t2} -like images at the
     two possible resolutions.
  5) The \regseg{} tool was applied using the warped test images as multispectral moving images and $\gammaset_R$ as shape priors.
  6) The agreement between the surfaces fitted by \regseg{} ($\hat{\gammaset}_{test}$) and $\gammaset_\text{true}$
     using automatically generated visual reports was assessed by computing the Hausdorff distance between the corresponding surfaces.}\label{fig:evphantoms}
\end{figure*}

\subsection{Real datasets} \label{sec:human_connectome}
The general experimental framework for the real datasets is presented in \autoref{fig:evworkflows},
which extends our previous evaluation \citep{esteban_simulationbased_2014} of distortions
  using \gls*{dmri} phantoms.

\begin{figure*}
\includegraphics[width=\linewidth]{figure03}
\caption{Experimental workflow employed to process real data from the \acrfull*{hcp}.
  1) $\gammaset_R$ were extracted from the anatomical reference (\gls*{t1} image).
  2) For use as the ground truth, we generated a plausible synthetic distortion $U_{true}$
    from the field map with \eqref{eq:fieldmap}.
  3) The \gls*{dmri} data were warped using $U_{true}$ to reproduce the effects of real
    susceptibility-derived distortions.
  Target diffusion scalars (\gls*{fa} and \gls*{adc}) were computed with the distorted data and
    stacked to feed the multivariate input required by \regseg{}.
  4) The method was run to obtain $U_{test} = \hat{U}_{true}$, i.e., the estimate of
    the ground-truth deformation.
  5) The results were evaluated visually and quantitatively.}\label{fig:evworkflows}
\end{figure*}

\paragraph*{Data}
To evaluate \regseg{} using real \gls*{dmri} data obtained from human brains,
  we collected 16 subjects from the \gls*{hcp} database.
The original acquisitions are released within ``unprocessed'' packages, whereas
  the ``minimally preprocessed'' are corrected for artifacts, brain-extracted,
  and spatially normalized, while some results have been subjected to the standard processing
  pipeline of \emph{FreeSurfer}.
We refer the reader to \citep{essen_human_2012} for exact details of the acquisition
  parameters and \citep{glasser_minimal_2013} for the preprocessing issues.
These datasets comprise a large set of images, including \gls*{t1}, \gls*{t2}, and
  multi-shell \gls*{dmri} images.

\begin{figure}
\includegraphics[width=\linewidth]{figure04}
\caption{Segmentation model defined by the homogeneous regions $\Omega_i$, where
  the image for segmentation is partitioned by the theoretical projection 
  ($\gammaset_{true}$) of $\gammaset_R$ onto $M$.
The plot represents a kernel density estimate of the location and spread of
  each region $\Omega_i$ in the bivariate feature space of $M$, which comprises 
  the \gls*{fa} and \gls*{adc} maps.
At the top and the right margins, the marginal distributions of $\Omega_i$ are also plotted for
  the \gls*{fa} and \gls*{adc}, respectively.
}\label{fig:model}
\end{figure}
